Syntax-Based Concept Alignment for Domain-Specific Machine Translation

### 2021-05-15 paper/paper.tex
added pseudocode + a not completely essential figure

@@

The pseudocode of the basic extraction algorithm is shown in Figure \ref{TODO:}\todo{add pseudocode}.
The input consists of a list of priority-sorted \textit{alignment criteria}, i.e. rules to determine whether two dependency trees should be aligned with each other which will be discussed in greater detail in Section \ref{criteria}, and a pair of sentences to align represented as UD trees.
From an implementation point of view, UD trees are, rose trees where each node represents a token, obtained from the CoNLL-U files produced by the UD parser.

\todo{add figure with the various tree representations?}

@@

Depending on which alignment criteria they match, the head of the two trees may or may not also be added to such collection. This is important not to miss one-word correspondences that cannot be captured in any other way, for instance between the root verbs of two full sentences.

@@

% TODO: pseudocode

>>>>

\usepackage{caption}
\usepackage{subcaption}

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

@@

The pseudocode of the basic extraction algorithm is shown below.

\begin{algorithm} 
  \small
  \begin{algorithmic}
  \Procedure{extract}{$criteria$,$(t,u)$}
    \State $alignments$ = $\emptyset$
    \If{$(t,u)$ matches any alignment $criteria$}
        \State $alignments$ += $(t,u)$
        \For{$(t',u')$ in \textsc{sort}(\textsc{subts}($t$)) $\times$ \textsc{sort}(\textsc{subts}($u$))}
          \State extract($criteria$,$(t',u')$)
        \EndFor
    \EndIf
    \Return $alignments$
  \EndProcedure
  \end{algorithmic}
  \end{algorithm}

The input consists of a list of priority-sorted \textit{alignment criteria}, i.e. rules to determine whether two dependency trees should be aligned with each other which will be discussed in greater detail in Section \ref{criteria}, and a pair of sentences to align represented as UD trees.
From an implementation point of view, UD trees are, rose trees where each node represents a token, obtained from the CoNLL-U files produced by the UD parser.

\begin{figure*}[h]
  \centering
  \begin{subfigure}{.23\textwidth}
      \centering
      \scriptsize
      \setlength{\unitlength}{0.16mm}
      \begin{picture}(371.0,130.0)
          \put(0.0,0.0){this}
          \put(46.0,0.0){is}
          \put(83.0,0.0){a}
          \put(120.0,0.0){dependency}
          \put(220.0,0.0){tree}
          \put(0.0,15.0){{\tiny PRON}}
          \put(46.0,15.0){{\tiny AUX}}
          \put(83.0,15.0){{\tiny DET}}
          \put(120.0,15.0){{\tiny NOUN}}
          \put(220.0,15.0){{\tiny NOUN}}
          \put(120.0,30.0){\oval(218.63636363636363,133.33333333333334)[t]}
          \put(10.681818181818187,35.0){\vector(0,-1){5.0}}
          \put(105.0,99.66666666666667){{\tiny nsubj}}
          \put(143.0,30.0){\oval(172.27586206896552,100.0)[t]}
          \put(56.86206896551724,35.0){\vector(0,-1){5.0}}
          \put(128.0,83.0){{\tiny cop}}
          \put(161.5,30.0){\oval(134.8102189781022,66.66666666666667)[t]}
          \put(94.0948905109489,35.0){\vector(0,-1){5.0}}
          \put(146.5,66.33333333333334){{\tiny det}}
          \put(180.0,30.0){\oval(97.0,33.333333333333336)[t]}
          \put(131.5,35.0){\vector(0,-1){5.0}}
          \put(165.0,49.66666666666667){{\tiny compound}}
          \put(235.0,130.0){\vector(0,-1){100.0}}
          \put(240.0,120.0){{\tiny root}}
      \end{picture}
  \end{subfigure}
  \begin{subfigure}{.38\textwidth}
    \centering
    \scriptsize
      \begin{verbatim}
      1	this	this	PRON	_	_	5	nsubj	_	_
      2	is	be	AUX	_	_	5	cop	_	_
      3	a	a	DET	_	_	5	det	_	_
      4	dependency	dependency	NOUN	_	_	5	
        compound	_	_
      5	tree	tree	NOUN	_	_	0	root	_	_
      \end{verbatim}
  \end{subfigure}%
  \begin{subfigure}{.38\textwidth}
    \centering
    \scriptsize
      \begin{verbatim}
5	tree	tree	NOUN	_	_	0	root	_	_
  4	dependency	dependency	NOUN	_	_	5	
    compound	_	_
  2	is	be	AUX	_	_	5	cop	_	_
  3	a	a	DET	_	_	5	det	_	_
  1	this	this	PRON	_	_	5	nsubj	_	_
      \end{verbatim}
  \end{subfigure}%
  \caption[The graphical representation of a UD tree alongside its CoNNL-U and sorted rose tree representation. Optional CoNNL-U fields are left blank.]{The graphical representation of a UD tree alongside its CoNNL-U and rose tree representation. Optional CoNNL-U fields are left blank.}
\end{figure*}

@@

@@

A simple but useful refinement is that, depending on which alignment criteria a pair of trees candidate for alignment matches, the heads of the two trees may or may not also be added to the collection of alignments. 
This is important not to miss one-word correspondences that cannot be captured in any other way, for instance between the root verbs of two full sentences.
A relevant implementative detail is that, in this context, the head of a tree is not simply defined as its root. Instead, if the root is part of a compound written as two or more separate words or a verb with auxiliaries, the root nodes of the corresponding children are also considered as part of its head.
